Expectations bias moral evaluations

### 2018-07-03 paper/paper.Rmd
updating pre-submission

significance: |
  People living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and thus are the very people in greatest need of assistance and concern from the world at-large. However, for these very reasons, it is often unsurprising when harm befalls them. Two experiments confirm this intuition: we found that people's expectations bias their reactions to morally-significant events--people are more upset by surprising harms than more expected harms. This bias threatens to reduce observer's concern for victims most in need, making people less likely to donate time or money to aid victims, or to take political action to prevent further harms.

@@

## change to true to add optional line numbering
lineno: TRUE

@@

When we learn of tragic events--a friend's loss of a loved-one, a violent killing, or a terrorist attack--we are mournful, concerned, angry, and driven to action. But our responses to these events differ depending on the context [e.g., @Trope2010]. In particular, our expectations play an important role in our reactions to events. This seems true of both the mundane--when a friend's loved-one passes, we often ask, "was it sudden?"--the dramatic--a murder in a sleepy town prompts a very different reaction than a similar crime in a major city--and even the catastrophic: acts of terrorism seem to evoke more outrage, horror, and concern when they occurs in peaceful rather than unstable regions. For instance, the 2015 Paris attacks were felt around the world. Yet, most of those mourning had little to say 15 hours earlier, when another tragic attack killed at least 43 people in Beirut [@Graham2015]. Likewise, stories of the Virginia Tech massacre, which killed 32 people, wildly overshadowed coverage of attacks killing nearly 200 people in Iraq--one of the bloodiest days since the 2003 invasion [@Greenslade2007]. In each of these situations, the surprise elicited by an event appears to affect people’s reactions to the event. In some contexts, harm is tragic and in others, “these things happen.”

Although this pattern seems so intuitive as to be obvious, no reasonable code of ethics would justify this pattern of judgment, nor do any moral psychological frameworks explicate the role of expectations in moral evaluations. Outside of the domain of morality, however, everyday experiences and controlled experiments show that people’s evaluations of events unambiguously depend on their expectations about those events. There is a special thrill to having one’s expectations exceeded and disappointment when events fail to meet expectations: we root for the underdog, hold surprise parties, and foreshadow bad news to ease its delivery [@Bell1985]. Indeed, in a controlled study, Mellers and colleagues [-@Mellers1997] found that expectations influenced affective reactions during a gambling task: Given a gamble with a 10% chance to win \$30 and 90% chance to win \$0, participants felt little disappointment with the \$0 outcome, but were considerably more excited when they won \$30. Conversely, given a gamble with a 90% chance of winning \$60 and 10% chance of winning \$0, participants were disappointed with the \$0 outcome and showed more muted enjoyment of the \$60 outcome. In fact, in gambles similar to these, people were happier with the smaller unexpected gain than with the larger but more expected gain [also see @Shepperd2002].

Applied to moral issues, the influences of expectations on evaluation would be a dangerous bias. Tragedies are tragic, whether expected or not, and failure to recognize that could lead to tacit endorsement of the pain and suffering of others wherever pain and suffering are the status quo. Do expectations directly bias reactions to morally consequential events, as intuition and anecdote seem to suggest? To demonstrate the plausibility of this suggestion, we develop a theoretical account of how expectations influence evaluations, not as a strange quirk or uniquely human bias, but instead as a fundamental component of how people evaluate events.

Most events of consequence consist of changes--a transition from one state to another. For this reason, we construe event evaluation as a comparison between the states of the world before and after the event. Comparison plays a crucial role in many theories of decision-making, and especially in theories aimed at explaining the role of expectations in evaluation  [e.g., @Bell1985; @Gul1991; @Loomes1986; @Mellers1997]. For instance, in Prospect Theory [@Kahneman1979; @Tversky1992], the utility of a prospect is evaluated by comparison to a current reference point. Our proposal is similar: we argue that expectations can set the reference points against which people compare future outcomes, so that assigning value to an event consists in comparing the state of the world following the event to the state of the world just prior to the event. As knowledge of the world is uncertain, our understanding of these states is probabilistic--that is, our model for the state of the world is just our expectations about what has happened, what is happening, and what will happen in the future. One direct implication of this proposal is that the evaluation of an event is linked to what is learned about the world as a result of that event’s occurrence. That is, we evaluate events positively or negatively to the extent that they inform our understanding of the world in positive or negative ways.

This account thereby provides an information-theoretic route to understanding how expectations directly influence evaluation. A fundamental insight of Information Theory is that the information carried by an event is a function of its prior probability; low probability events carry more information than high probability events [@Shannon1948]. We propose that people learn more about the state of the world when their expectations are violated by shocking world events as compared to when they are affirmed by less surprising events. For example, if a bombing occurs in Paris--a less expected location--rather than Lebanon--a more expected location--we learn that the world is more dangerous than we had previously believed. This suggestion is novel in the moral domain, but these points are widely accepted elsewhere. For instance, expectations have long been recognized as fundamental to models of human and animal learning [e.g., @Rescorla1972]. Similarly, science progresses most when new findings impugn widely-accepted theories or when a theory’s extreme predictions are validated.

@@

\begin{equation}
E_X[u(X)] = \sum_x u(x)p(x)
\end{equation}

@@

\begin{equation}
E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
\end{equation}

Given our construal of event evaluation as the comparison between states of the world following and prior to the event, we define the value assigned to the signalling event, $V(S)$, as:

@@

@@

The design of these studies was chosen with special consideration to two theoretical points. First, the information-theoretic account we have described makes a solely directional prediction--all else being equal, more surprising events will elicit stronger reactions than less surprising events. Asking participants to compare two similar events allowed us to test this directional prediction. Second, it is the informativeness of events, determined by one's prior expectations of the event, that we predict will shape evaluations. Thus, reasoners should not have to be induced to hold expectations experimentally [as in experiments by @Mellers1997], but instead their own prior beliefs will be sufficient to shape their expectations and affect their evaluations. For this reason, we manipulated the prior probability of events by describing two otherwise identical events occurring in different contexts. This allowed us to test the role of expectations based on world knowledge, as well as providing a more naturalistic set of items.

@@

Consistent with our predictions, manipulating expectations consistently shaped evaluations across a range of morally-significant events. We found that people viewed unexpected harmful events as more upsetting than expected harmful events, even when the harm a victim suffered was the same. For instance, people judged that a robbery at a clothing store (unexpected) was more upsetting than a robbery at a convenience store (expected). Furthermore, these differences were large, systematic, and robust. Although there was variability in the situations that elicited the largest differences, our statistical approach treated trial as a random effect (see Methods and Results), allowing us to generalize conclusions to items we did not test [@Judd2012].

@@

Yet, however well-grounded the underlying principles, there’s a clear moral danger whenever expectations influence our moral concerns. For instance, "victim blaming" describes the tendency to shift blame away from aggressors and onto victims of crimes like sexual assault [@Koss2000] and police violence. Work by Janoff-Bulman and colleagues [-@Janoff-bulman1985] has shown that merely increasing the perceived probability of a victim suffering a sexual assault induced participants to blame the victim for her assault. People’s expectations may explain the flip side of victim blaming, whereby those who victim blame also tend to judge transgressors less harshly: When people perceive a victims’ actions as increasing the probability of their eventual assault, they may perceive an aggressor’s actions as producing a smaller change in utility, thus warranting weaker condemnation.

The influence of expectations on moral evaluations poses perhaps the greatest threat when considering our reactions to humanitarian challenges brought on by poverty or regional instability. In these cases, the influence of expectations on moral evaluations threatens to impose vicious and morally pernicious cycles. For instance, people living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and are the very people in greatest need of assistance and concern from the world at-large. However, for these very reasons, it is often unsurprising when harm befalls people living in these circumstances. Thus the influence of expectations on moral evaluations threatens to reduce observer's concern for these victims most in need, making people less likely to donate time or money to aid victims, or to take political action to prevent further harms. The present research identifies this potentially pernicious moral bias and grounds it in information theoretic principles. Hopefully, by further understanding these cognitive dynamics, psychologists and policymakers can begin to find ways to reduce or counteract people's tendency to ignore the plight of those in most need of attention.

@@

We conducted two preregistered studies examining the role of expectations in moral evaluations using two different judgment tasks (https://osf.io/5qnf7/). Each study consisted of multiple trials where participants were presented brief descriptions of two events. Their task was to decide which of the two events was more upsetting. In Study 1, they made a two-alternative forced-choice, and in Study 2 they made their judgments using a five-point rating scale.  Both studies consisted of 16 experimental trials, five “equivalent” filler trials, and five “non-equivalent” filler trials.

@@

Both studies also included “equivalent” filler trials. In these trials, the two events differed in trivial contextual details that we did not expect would affect participants’ judgments. These filler trials were meant to prevent participants from becoming explicitly aware of the structure of the experimental trials. Finally, both studies included “non-equivalent” filler trials, the two events differed substantially in the degree of harm suffered by a victim, so that one event was expected to be seen as considerably more upsetting than the other. These trials were included to allow participants a chance to use the extremes of the response scale and to reduce any task demands that might drive them to make artificially fine-grained distinctions when generating their responses. All experimental materials for these studies are available as Supporting Information at https://osf.io/5qnf7/.

@@

We note that the most common response was to view both events, which involve identical harms, as equivalently upsetting. However, when participants did view one of the events as more upsetting, they showed a clear bias towards choosing unexpected events as more upsetting. Thus, Study 2 replicates the tendency to view unexpected moral events as more upsetting using a more conservative judgment task.

@@

Evaluation (e.g., of utility) is often examined in the context of decision-making. However, people often generate post hoc evaluations of events for their own sake. For instance, to reflect on the results of our choices and to evaluate others--to judge the wisdom of their decisions or the morality of their actions.

@@

\begin{equation}
E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
\end{equation}

@@

Applying the expectation of conditional expectation, we can write:

\begin{equation}
V(s=1) = \sum_x u(x)p(x|s=1) - \sum_x \sum_s u(x)p(x|s)p(s)
\end{equation}

We then perform some algebra. We break out the summation over $S$ (6), pull the multipliers out of the last two sums (7), rewrite $p(s=0)$ in terms of $p(s=1)$ (8), and factor to combine the first two sums (9).

\begin{align}
V(s=1) &= \sum_x u(x)p(x|s=1) - \sum_x u(x)p(x|s=1)p(s=1) - \sum_x u(x)p(x|s=0)p(s=0) \\
 &= \sum_x u(x)p(x|s=1) - p(s=1)\sum_x u(x)p(x|s=1) - p(s=0) \sum_x u(x)p(x|s=0) \\
 &= \sum_x u(x)p(x|s=1) - p(s=1)\sum_x u(x)p(x|s=1) - (1-p(s=1)) \sum_x u(x)p(x|s=0) \\
 &=(1-p(s=1))\sum_x u(x)p(x|s=1) - (1-p(s=1)) \sum_x u(x)p(x|s=0)
\end{align}

Finally, combining both sums by their common factor yields (10):

@@

Thus, it can be seen under this formalization that the role of expectations in the evaluation of events is fundamental--it does not, for instance, depend on any specifics concerning the utility function over states of the world. Instead, this conclusion depends on just two assumptions: first, that events are evaluated by comparing the utilities of the states of the world prior to and following the event; and second, that knowledge of those states of the world is uncertain.

We have appealed to the informativeness of an event to illustrate intuitively how expectations should influence evaluations. More expected events carry less information, and thereby engender weaker reactions. Though our model of event evaluation is not an information-theoretic model in a direct sense, this interpretation still applies: the information carried by observing $S$ is a function of that observation’s prior probability, which we have in turn shown determines its evaluation. Also relevant, although not addressed here, is the mutual information between $X$ and $S$--the extent to which $S$ informs $X$.

>>>>

significance: |
  In two behavioral studies, we found that people's expectations biased their reactions to morally-significant events--people are more upset by surprising harms than more expected harms. This bias appears deeply pernicious: because it is often unsurprising when people living in poverty, dangerous neighborhoods, or unstable regions become victims of crime, natural disasters, or terrorism, observers are likely to experience reduced concern for these victims. Yet people living in disadvantaged circumstances are those most in need of assistance and concern from the world at-large. By blunting people's concern for those victims who are most in need, the bias we have uncovered threatens to increase their suffering and further sustain inequality.

@@

## change to true to add optional line numbering
lineno: FALSE

@@

When we learn of tragic events--a friend's loss of a loved-one, a violent killing, or a terrorist attack--we are mournful, concerned, angry, and driven to action. But our responses to these events differ depending on the context in which these events occur [e.g., @Trope2010]. In particular, our expectations play an important role in our reactions to events. This seems true of both the mundane--when a friend's loved-one passes, we often ask, "was it sudden?"--the dramatic--a murder in a sleepy town prompts a very different reaction than a similar crime in a major city--and even the catastrophic: acts of terrorism seem to evoke more outrage, horror, and concern when they occurs in peaceful rather than unstable regions. For instance, the 2015 Paris attacks were felt around the world. Yet, most of those mourning had little to say 15 hours earlier, when another tragic attack killed at least 43 people in Beirut [@Graham2015]. Likewise, stories of the Virginia Tech massacre, which killed 32 people, wildly overshadowed coverage of attacks killing nearly 200 people in Iraq--one of the bloodiest days since the 2003 invasion [@Greenslade2007]. In each of these situations, the surprise elicited by an event appears to affect people’s reactions to the event. In some contexts, harm is tragic and in others, “these things happen.”

Although this pattern seems so intuitive as to be obvious, no reasonable code of ethics would justify this pattern of judgment, nor do any moral psychological frameworks explicate the role of expectations in moral evaluations. Outside of the domain of morality, everyday experiences and controlled experiments show that people’s evaluations of events unambiguously depend on their expectations about those events. There is a special thrill to having one’s expectations exceeded and disappointment when events fail to meet expectations: we root for the underdog, hold surprise parties, and foreshadow bad news to ease its delivery [@Bell1985]. Indeed, in a controlled study, Mellers and colleagues [-@Mellers1997] found that expectations influenced affective reactions during a gambling task: Given a gamble with a 10% chance to win \$30 and 90% chance to win \$0, participants felt little disappointment with the \$0 outcome, but were considerably more excited when they won \$30. Conversely, given a gamble with a 90% chance of winning \$60 and 10% chance of winning \$0, participants were disappointed with the \$0 outcome and showed more muted enjoyment of the \$60 outcome. In fact, in gambles similar to these, people were happier with the smaller unexpected gain than with the larger but more expected gain [also see @Shepperd2002].

The influence of expectations on moral evaluations threatens to produce negative social consequences. Tragedies are tragic, whether expected or not, and failure to recognize that could lead to tacit endorsement of the pain and suffering of others wherever pain and suffering are the status quo. Do expectations directly bias reactions to morally consequential events, as intuition and anecdote seem to suggest? To demonstrate the plausibility of this suggestion, we develop a theoretical account of how expectations influence evaluations, not as a strange quirk or uniquely human bias, but instead as a fundamental component of how people evaluate events.

Most events of consequence consist of changes--a transition from one state to another. For this reason, we construe event evaluation as a comparison between the states of the world before and after the event. Comparison plays a crucial role in many theories of decision-making, and especially in theories aimed at explaining the role of expectations in evaluation  [e.g., @Bell1985; @Gul1991; @Loomes1986; @Mellers1997]. For instance, in Prospect Theory [@Kahneman1979; @Tversky1992], the utility of a prospect is evaluated by comparison to a current reference point. Our proposal is similar: we argue that expectations can set the reference points against which people compare future outcomes, so that assigning value to an event consists in comparing the state of the world following the event to the state of the world just prior to the event. We further assume that people’s knowledge of the world is uncertain, so we represent the states of the world probabilistically. This means that a person’s model for the state of the world is just their expectations about what has happened, what is happening, and what will happen in the future. One direct implication of this proposal is that the evaluation of an event is linked to what is learned about the world as a result of that event’s occurrence. That is, we evaluate events positively or negatively to the extent that they inform our understanding of the world in positive or negative ways.

One striking implication of this account is that it provides an information-theoretic route to understanding how expectations directly influence evaluation. A fundamental insight of Information Theory is that the information carried by an event is a function of its prior probability; low probability events carry more information than high probability events [@Shannon1948]. We propose that people learn more about the state of the world when their expectations are violated by shocking world events as compared to when they are affirmed by less surprising events. For example, if a bombing occurs in Paris--a less expected location--rather than Lebanon--a more expected location--we learn that the world is more dangerous than we had previously believed. This suggestion is novel in the moral domain, but these points are widely accepted elsewhere. For instance, expectations have long been recognized as fundamental to models of human and animal learning [e.g., @Rescorla1972].

@@

\begin{equation*}
E_X[u(X)] = \sum_x u(x)p(x)
\end{equation*}

@@

\begin{equation*}
E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
\end{equation*}

Our construal of event evaluation is as a comparison between states of the world following and prior to an event. Consequently, we define the value assigned to the signalling event, $V(S)$, as:

\begin{equation*}
V(S) = E_{X|S}[u(X)] - E_X[u(X)]
\end{equation*}

@@

\begin{equation*}
V(s=1)=(1-p(s=1))\sum_x u(x)p(x|s=1) - u(x)p(x|s=0)
\end{equation*}

@@

The design of these studies was chosen with special consideration to two theoretical points. First, the information-theoretic account we have described makes a solely directional prediction--all else being equal, more surprising events will elicit stronger reactions than less surprising events. Asking participants to compare two similar events allowed us to test this directional prediction. Second, it is the informativeness of events, determined by one's prior expectations of the event, that we predict will shape evaluations. Thus, reasoners should not have to be induced to hold expectations experimentally [as in experiments by @Mellers1997], but instead their own prior beliefs will be sufficient to shape their expectations and affect their evaluations. For this reason, we manipulated the prior probability of events by describing two otherwise identical events occurring in different contexts. We explicitly chose contexts that differed in ways that could not be accounted for by construal-level explanations (@Trope2010) or explanations based on biases towards one’s ingroup (@Brewer1979). This allowed us to test the role of expectations based on world knowledge, as well as providing a more naturalistic set of items.

@@

Consistent with our predictions, manipulating expectations consistently shaped evaluations across a range of morally-significant events. We found that people viewed unexpected harmful events as more upsetting than expected harmful events, even when the harm a victim suffered was the same. For instance, people judged that a robbery at a clothing store (unexpected) was more upsetting than a robbery at a convenience store (expected). Furthermore, these differences were large, systematic, and robust. Although there was variability in the situations that elicited the largest differences, our statistical approach treated trial as a random effect (see Methods and Results), allowing us to generalize our conclusions to items we did not test [@Judd2012].

@@

Yet, however well-grounded the underlying principles, there’s a clear moral danger whenever expectations influence our moral concerns [@Koss2000; -@Janoff-bulman1985]. These effects pose a significant threat when considering our reactions to humanitarian challenges brought on by poverty or regional instability. In these cases, the influence of expectations on moral evaluations threatens to impose vicious and morally pernicious cycles. For instance, people living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and are the very people in greatest need of assistance and concern from the world at-large. However, for these very reasons, it is often unsurprising when harm befalls people living in these circumstances. Thus the influence of expectations on moral evaluations threatens to reduce observer's concern for these victims most in need, making people less likely to donate time or money to aid victims, or to take political action to prevent further harms. The present research identifies this potentially pernicious moral bias and grounds it in information theoretic principles. Hopefully, by further understanding these cognitive dynamics, psychologists and policymakers can begin to find ways to reduce or counteract people's tendency to ignore the plight of those in most need of attention.

@@

We conducted two preregistered studies examining the role of expectations in moral evaluations (https://osf.io/86rsw/). Each study consisted of multiple trials where participants were presented brief descriptions of two events. Their task was to decide which of the two events was more upsetting. In Study 1, they made a two-alternative forced-choice, and in Study 2 they made their judgments using a five-point rating scale.  Both studies consisted of 16 experimental trials, five “equivalent” filler trials, and five “non-equivalent” filler trials.

@@

Both studies also included “equivalent” filler trials. In these trials, the two events differed in trivial contextual details that we did not expect would affect participants’ judgments. These filler trials were meant to prevent participants from becoming explicitly aware of the structure of the experimental trials. Finally, both studies included “non-equivalent” filler trials, the two events differed substantially in the degree of harm suffered by a victim, so that one event was expected to be seen as considerably more upsetting than the other. These trials were included to allow participants a chance to use the extremes of the response scale and to reduce any task demands that might drive them to make artificially fine-grained distinctions when generating their responses. All experimental materials for these studies are available as Supporting Information at https://osf.io/86rsw/.

@@

We note that the most common response was to view both events, which involve identical harms, as equivalently upsetting. However, across almost every item, participants showed a clear bias towards choosing unexpected events as more upsetting. Thus, Study 2 replicates the tendency to view unexpected moral events as more upsetting using a more conservative judgment task.

@@

Evaluation of utility is often examined in the context of decision-making. However, people often generate post hoc evaluations of events for their own sake. For instance, people evaluate events to judge the wisdom of their decisions or the morality of their actions.

@@

\begin{equation*}
E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
\end{equation*}

@@

Our goal is to express this relation in terms of $p(S)$. First, we manipulate $E_X[u(X)]$ by applying the expectation of conditional expectation and some algebra.

\begin{align*}
\sum_x u(x)p(x) & = \sum_s u(x)p(x|s)p(s) \\
 &= \begin{aligned}[t]
    & \sum_x u(x)p(x|s=1)p(s=1) - \\
    & \quad \sum_x u(x)p(x|s=0)p(s=0) \end{aligned}\\
 &= \begin{aligned}[t]
    & p(s=1)\sum_x u(x)p(x|s=1) -\\
    & \quad p(s=0) \sum_x u(x)p(x|s=0) \end{aligned}\\
 &= \begin{aligned}[t]
    & p(s=1)\sum_x u(x)p(x|s=1) - \\
    & \quad (1-p(s=1)) \sum_x u(x)p(x|s=0) \end{aligned}\\
\end{align*}

Substituting this quantity back into Equation 2, we obtain:

\begin{align*}
V(s=1) = & \sum_x u(x)p(x|s=1) - \\ & p(s=1)\sum_x u(x)p(x|s=1) -\\ & (1-p(s=1)) \sum_x u(x)p(x|s=0)
\end{align*}

Finally, we factor the first two sums and combine them with the last sum to yield:

@@

Thus, it can be seen under this formalization that the role of expectations in the evaluation of events is fundamental in the sense that it does not depend on any specifics concerning the utility function over states of the world. Instead, this conclusion depends on just two assumptions: first, that events are evaluated by comparing the utilities of the states of the world prior to and following the event; and second, that knowledge of those states of the world is uncertain.

We have appealed to the informativeness of an event to illustrate intuitively how expectations should influence evaluations. More expected events carry less information, and thereby engender weaker reactions. Though our model of event evaluation is not an information-theoretic model in a direct sense, this interpretation still applies: the information carried by observing $S$ is a function of that observation’s prior probability, which we have in turn shown determines its evaluation.
